---
title: "Introduction to R Notebooks"
output: html_notebook
---

# Getting Started with R Notebooks

R Notebooks are interactive documents that combine code, output, and narrative text. They're perfect for data analysis, teaching, and sharing reproducible research.

## What Makes R Notebooks Special?

Unlike traditional R scripts, notebooks:
- Display code output inline
- Support rich text formatting with Markdown
- Allow for interactive exploration
- Can be executed chunk by chunk
- Export to HTML, PDF, or Word formats

Let's explore the key features!

## Creating and Running Code Chunks

Code chunks are enclosed in triple backticks with `{r}` to indicate R code:

```{r}
# This is an R code chunk
x <- 5
y <- 10
x + y
```

The output appears directly below the code. To run a chunk:
- Click the green "play" button in the top-right corner of the chunk
- Use the keyboard shortcut Ctrl+Shift+Enter (Cmd+Shift+Enter on Mac)
- To run a single line, position your cursor on that line and press Ctrl+Enter

## Chunk Options

You can customize how chunks behave using options:

```{r, echo=FALSE}
# This code runs but isn't displayed in the output
secret_calculation <- 123 * 456
```

```{r, eval=FALSE}
# This code is displayed but not executed
this_wont_run <- "example"
```

Common chunk options:
- `echo`: Whether to display the code (TRUE/FALSE)
- `eval`: Whether to evaluate/run the code (TRUE/FALSE)
- `warning`: Whether to show warnings (TRUE/FALSE)
- `message`: Whether to show messages (TRUE/FALSE)
- `fig.width`, `fig.height`: Control the size of figures

## Creating Visualizations

Visualizations render directly in the notebook:

```{r}
# Basic plot example
plot(cars, main = "Speed vs. Stopping Distance", 
     xlab = "Speed (mph)", ylab = "Stopping Distance (ft)")
```

And here's a more advanced example with ggplot2:

```{r}
# Load ggplot2 package
library(ggplot2)

# Create a nicer visualization
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Relationship Between Speed and Stopping Distance",
       x = "Speed (mph)",
       y = "Stopping Distance (ft)") +
  theme_minimal()
```

## Inline R Code

You can include R output directly within text using backticks with `r`:

The average stopping distance in the cars dataset is `r round(mean(cars$dist), 1)` feet, and the average speed is `r round(mean(cars$speed), 1)` mph.

## Interactive Tables

Tables are a great way to display data:

```{r}
# Basic table
head(mtcars[, 1:5])
```

For more attractive tables, try the kable function from knitr:

```{r}
library(knitr)
kable(head(mtcars[, 1:5]), caption = "First 6 rows of mtcars dataset")
```

## Working with Data

Here's a simple data analysis workflow:

```{r}
# Load and examine data
data(iris)
str(iris)

# Basic summary statistics
summary(iris$Sepal.Length)

# Group-level statistics
aggregate(Sepal.Length ~ Species, data = iris, FUN = function(x) {
  c(mean = mean(x), sd = sd(x), min = min(x), max = max(x))
})
```

## Mathematical Expressions and LaTeX

R Notebooks fully support LaTeX for mathematical notation, making them ideal for statistical, scientific, and mathematical content. LaTeX can be used in two ways:

### Inline Math Expressions

For inline equations, use single dollar signs `---
title: "Introduction to R Notebooks"
output: html_notebook
---

# Getting Started with R Notebooks

R Notebooks are interactive documents that combine code, output, and narrative text. They're perfect for data analysis, teaching, and sharing reproducible research.

## What Makes R Notebooks Special?

Unlike traditional R scripts, notebooks:
- Display code output inline
- Support rich text formatting with Markdown
- Allow for interactive exploration
- Can be executed chunk by chunk
- Export to HTML, PDF, or Word formats

Let's explore the key features!

## Creating and Running Code Chunks

Code chunks are enclosed in triple backticks with `{r}` to indicate R code:

```{r}
# This is an R code chunk
x <- 5
y <- 10
x + y
```

The output appears directly below the code. To run a chunk:
- Click the green "play" button in the top-right corner of the chunk
- Use the keyboard shortcut Ctrl+Shift+Enter (Cmd+Shift+Enter on Mac)
- To run a single line, position your cursor on that line and press Ctrl+Enter

## Chunk Options

You can customize how chunks behave using options:

```{r, echo=FALSE}
# This code runs but isn't displayed in the output
secret_calculation <- 123 * 456
```

```{r, eval=FALSE}
# This code is displayed but not executed
this_wont_run <- "example"
```

Common chunk options:
- `echo`: Whether to display the code (TRUE/FALSE)
- `eval`: Whether to evaluate/run the code (TRUE/FALSE)
- `warning`: Whether to show warnings (TRUE/FALSE)
- `message`: Whether to show messages (TRUE/FALSE)
- `fig.width`, `fig.height`: Control the size of figures

## Creating Visualizations

Visualizations render directly in the notebook:

```{r}
# Basic plot example
plot(cars, main = "Speed vs. Stopping Distance", 
     xlab = "Speed (mph)", ylab = "Stopping Distance (ft)")
```

And here's a more advanced example with ggplot2:

```{r}
# Load ggplot2 package
library(ggplot2)

# Create a nicer visualization
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Relationship Between Speed and Stopping Distance",
       x = "Speed (mph)",
       y = "Stopping Distance (ft)") +
  theme_minimal()
```

## Inline R Code

You can include R output directly within text using backticks with `r`:

The average stopping distance in the cars dataset is `r round(mean(cars$dist), 1)` feet, and the average speed is `r round(mean(cars$speed), 1)` mph.

## Interactive Tables

Tables are a great way to display data:

```{r}
# Basic table
head(mtcars[, 1:5])
```

For more attractive tables, try the kable function from knitr:

```{r}
library(knitr)
kable(head(mtcars[, 1:5]), caption = "First 6 rows of mtcars dataset")
```

## Working with Data

Here's a simple data analysis workflow:

```{r}
# Load and examine data
data(iris)
str(iris)

# Basic summary statistics
summary(iris$Sepal.Length)

# Group-level statistics
aggregate(Sepal.Length ~ Species, data = iris, FUN = function(x) {
  c(mean = mean(x), sd = sd(x), min = min(x), max = max(x))
})
```

:

This regression has a slope of $\beta_1 = 2.59$ and an R-squared of $R^2 = 0.75$.

### Display Equations

For standalone equations, use double dollar signs `$`:

The formula for a simple linear regression is:

$\hat{y} = \beta_0 + \beta_1 x$

Where:
- $\hat{y}$ is the predicted value
- $\beta_0$ is the intercept
- $\beta_1$ is the slope coefficient
- $x$ is the predictor variable

### Multi-line Equations

You can create aligned multi-line equations:

$\begin{align}
\log\left(\frac{p(x)}{1-p(x)}\right) &= \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \ldots + \beta_p x_p \\
&= \mathbf{X}\boldsymbol{\beta}
\end{align}$

### Statistical Notation

LaTeX is particularly useful for statistical notation:

The t-test statistic is calculated as:

$t = \frac{\bar{x} - \mu}{s / \sqrt{n}}$

And the F-statistic for ANOVA is:

$F = \frac{\text{MSB}}{\text{MSW}} = \frac{\text{Between-group variability}}{\text{Within-group variability}}$

### Matrices and Arrays

You can typeset matrices and arrays:

$\mathbf{X} = 
\begin{pmatrix} 
1 & x_{11} & x_{12} & \ldots & x_{1p} \\
1 & x_{21} & x_{22} & \ldots & x_{2p} \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
1 & x_{n1} & x_{n2} & \ldots & x_{np}
\end{pmatrix}$

### Combining LaTeX with R Output

You can combine LaTeX with inline R code:

The model coefficients are $\hat{\beta}_0 = `r coef(model)[1]`$ and $\hat{\beta}_1 = `r coef(model)[2]`$.

## Organizing with Headers and Text

Use Markdown to structure your notebook:

# Level 1 Header
## Level 2 Header
### Level 3 Header

You can also use:
- Bullet lists
- *Italic text*
- **Bold text**
- [Links](https://www.rstudio.com)
- Horizontal rules like this:

---

## Warning and Error Messages

Error messages and warnings are displayed inline:

```{r}
# This will show a warning
log(-1)

# This would show an error if not wrapped in tryCatch
tryCatch(
  1/0,
  error = function(e) paste("Error caught:", e$message)
)
```

## Caching for Time-Consuming Computations

For long-running code, use caching to avoid re-computation:

```{r, cache=TRUE}
# This result will be cached
Sys.sleep(2)  # Simulating long computation
result <- "Computation complete"
result
```

## Including External Files

You can read external files and display their content:

```{r, eval=FALSE}
# Reading a CSV file (example, will throw an error without an actual file)
my_data <- read.csv("data.csv")
head(my_data)
```

## Tips for Effective R Notebooks

1. **Start with an outline**: Use headers to structure your analysis
2. **Mix code with narrative**: Explain your approach and findings
3. **Keep code chunks focused**: Each chunk should do one clear task
4. **Use inline code** for simple values to maintain consistency
5. **Add comments to complex code** for future readers
6. **Include visualizations** to illustrate your findings
7. **Cache time-consuming operations** with `cache=TRUE`
8. **Preview regularly** to check formatting and output

## Working with R Notebooks in RStudio

### Preview and Publishing

RStudio provides excellent integration with R Notebooks:

1. **Preview Feature**: Click "Preview" (or press Ctrl+Shift+K) to generate a `.nb.html` file
   - This HTML file contains both the code and output
   - It preserves the interactive nature of your notebook
   - Viewers can see your code, outputs, and narrative without running R

2. **Publishing Workflow**:
   - Preview generates `your-notebook.nb.html` alongside your `your-notebook.Rmd` file
   - The `.nb.html` file can be shared via email, web servers, or platforms like RPubs
   - For formal reports, use "Knit to HTML/PDF/Word" for more polished output

### Best Practices for Version Control

When using version control (Git/GitHub) with R Notebooks:

1. **Commit the .Rmd file**: Always commit your source R Markdown file
   
2. **Running code before committing**:
   - Generally, commit notebooks with code that has been run and verified
   - This ensures the notebook contains valid code and expected outputs
   - Exception: For very large outputs or time-consuming computations, you might commit without running

3. **.gitignore recommendations**:
   - Consider adding `.nb.html` files to `.gitignore` if they're large or change frequently
   - Alternatively, commit them if you want a record of outputs or for others to view without running
   
4. **Cache directories**:
   - Add `*_cache/` and `*_files/` to `.gitignore` to avoid committing cached results

## Architecture for Notebook Projects

For larger projects with multiple notebooks:

1. **Directory Structure**:
   ```
   project/
   ├── data/               # Raw data files
   ├── processed_data/     # Cleaned/transformed data
   ├── notebooks/          # R notebooks for analysis
   ├── R/                  # Reusable R functions
   ├── output/             # Generated figures and tables
   └── README.md           # Project documentation
   ```

2. **Code Organization**:
   - Keep data cleaning operations in separate notebooks or scripts
   - Use source() to load common functions from the R/ directory
   - Use relative paths with here::here() for file references

3. **Setup Chunks**:
   Include a setup chunk at the beginning to load libraries and source files:

```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load essential libraries
library(dplyr)      # For data manipulation
library(ggplot2)    # For data visualization
library(readr)      # For reading data files
library(knitr)      # For tables and output formatting

# If you have custom helper functions, source them
# source("R/helper_functions.R")

# Set default ggplot theme
theme_set(theme_minimal())

# For reproducibility
set.seed(123)
```

- See the notebook on random number reproducibility in R for more explanation

4. **Data Handling**:
```{r load-data, include=FALSE}
   # Load and process data without showing verbose output
   data <- read.csv(here("data", "raw_data.csv"))
```

5. **Output Management**:
```{r save-results, include=FALSE}
   # Save results without showing code
   ggsave(here("output", "figure1.png"), plot = my_plot)
   write.csv(results, here("output", "results.csv"))
```

## Keyboard Shortcuts

Useful keyboard shortcuts for R Notebooks:
- Insert code chunk: Ctrl+Alt+I (Cmd+Option+I on Mac)
- Run current chunk: Ctrl+Shift+Enter
- Run current line: Ctrl+Enter
- Run all chunks above: Alt+Ctrl+P
- Run all chunks: Ctrl+Alt+R

## Interactive Example: Linear Regression

Let's pull everything together with a complete analysis example:

```{r}
# Load data
data(mtcars)

# Explore relationship
plot(mtcars$wt, mtcars$mpg, 
     main = "Car Weight vs. Fuel Efficiency",
     xlab = "Weight (1000 lbs)", 
     ylab = "Miles Per Gallon",
     pch = 19, col = "darkblue")

# Fit linear model
model <- lm(mpg ~ wt, data = mtcars)

# Add regression line
abline(model, col = "red", lwd = 2)

# Model summary
summary(model)

# Diagnostic plots
par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))
```

## Additional Best Practices

### Managing Verbose Output

For operations that generate verbose output (like data cleaning):

```{r message=FALSE, warning=FALSE, results='hide'}
# This code will run but won't display messages, warnings, or results
library(dplyr)
large_dataset <- read.csv("large_file.csv") %>%
  filter(!is.na(important_column)) %>%
  mutate(new_var = complicated_calculation(old_var))
```

### Documentation vs. Analysis Notebooks

Consider creating different types of notebooks:

1. **Processing notebooks**: Focus on data cleaning and preparation
   - More code-heavy
   - Include detailed comments
   - Often run once to generate clean data

2. **Analysis notebooks**: Focus on insights and visualization
   - More narrative-focused
   - Include interpretations of results
   - Designed for sharing with stakeholders

3. **Report notebooks**: Polished presentation of findings
   - Minimal visible code (`echo=FALSE` for most chunks)
   - Emphasis on visualizations and conclusions
   - Formatted for non-technical audiences

### Interactive Elements

For complex projects, consider adding interactive elements:

```{r eval=FALSE}
# Adding interactive widgets
library(shiny)
library(plotly)

# Create interactive plot
p <- ggplotly(ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point())
p
```

## Conclusion

R Notebooks provide a powerful way to combine code, output, and narrative in a single document. They're perfect for:
- Data analysis reports
- Teaching materials
- Research papers
- Sharing reproducible work

With proper organization and best practices, R Notebooks can streamline your workflow from data cleaning to publication-ready reports, while maintaining reproducibility throughout the process.

Now you're ready to create your own R Notebooks!