---
title: "Company Salary Distribution Analysis"
author: "Data Analyst"
date: "14 March 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Check and install required packages if needed
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("dplyr")) install.packages("dplyr")
if (!require("scales")) install.packages("scales")
if (!require("knitr")) install.packages("knitr")
if (!require("tidyr")) install.packages("tidyr")

# Load required libraries
library(ggplot2)
library(dplyr)
library(scales)
library(knitr)
library(tidyr)
```

# Salary Distribution Analysis

## Company Parameters

```{r company_params}
# Company parameters
total_monthly_salary <- 461000  # Total monthly salary expenditure in GBP
num_employees <- 70             # Number of employees
mean_salary <- total_monthly_salary / num_employees  # Mean monthly salary

# Display basic statistics
company_stats <- data.frame(
  Statistic = c("Total Monthly Salary Expenditure", "Number of Employees", "Mean Monthly Salary"),
  Value = c(paste0("£", format(total_monthly_salary, big.mark=",")), 
            num_employees, 
            paste0("£", format(mean_salary, digits=2, big.mark=",", nsmall=2)))
)

kable(company_stats, align = "lr", caption = "Company Parameters")
```

## Log-normal Distribution Parameters

We model the salary distribution using a log-normal distribution, which is commonly used for modelling salary data due to its right-skewed nature.

```{r lognormal_params}
# Assume a coefficient of variation (CV) for salaries
cv <- 0.5  # Typical for medium-sized companies

# Calculate log-normal parameters
sigma_squared <- log(1 + cv^2)
sigma <- sqrt(sigma_squared)
mu <- log(mean_salary) - sigma_squared/2

# Display log-normal parameters
lognormal_params <- data.frame(
  Parameter = c("μ (mu)", "σ (sigma)", "σ² (sigma squared)"),
  Value = c(round(mu, 4), round(sigma, 4), round(sigma_squared, 4)),
  Description = c("Mean of logarithm of salary", 
                  "Standard deviation of logarithm of salary", 
                  "Variance of logarithm of salary")
)

kable(lognormal_params, align = "llr", caption = "Log-normal Distribution Parameters")
```

## Visualising the Salary Distribution

### Probability Density Function

```{r pdf_plot, fig.width=10, fig.height=6}
# Create sequence of salary values
salary_seq <- seq(1000, 25000, by = 100)

# Calculate PDF values
pdf_values <- dlnorm(salary_seq, meanlog = mu, sdlog = sigma)

# Create data frame for plotting
pdf_df <- data.frame(
  Salary = salary_seq,
  Density = pdf_values
)

# Plot PDF
pdf_plot <- ggplot(pdf_df, aes(x = Salary, y = Density)) +
  geom_line(size = 1, color = "steelblue") +
  geom_vline(xintercept = mean_salary, linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = exp(mu), linetype = "dotted", color = "darkgreen", size = 1) +
  annotate("text", x = mean_salary + 1500, y = max(pdf_values) * 0.9, 
           label = "Mean", color = "red") +
  annotate("text", x = exp(mu) - 1500, y = max(pdf_values) * 0.9, 
           label = "Median", color = "darkgreen") +
  labs(
    title = "Log-normal Probability Density Function of Salary Distribution",
    x = "Monthly Salary (£)",
    y = "Density"
  ) +
  scale_x_continuous(labels = scales::comma_format(prefix = "£")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

print(pdf_plot)
```

### Cumulative Distribution Function

```{r cdf_plot, fig.width=10, fig.height=6}
# Calculate CDF values
cdf_values <- plnorm(salary_seq, meanlog = mu, sdlog = sigma)

# Create data frame for plotting
cdf_df <- data.frame(
  Salary = salary_seq,
  Cumulative_Probability = cdf_values
)

# Plot CDF
cdf_plot <- ggplot(cdf_df, aes(x = Salary, y = Cumulative_Probability)) +
  geom_line(size = 1, color = "darkred") +
  labs(
    title = "Cumulative Distribution Function of Salary Distribution",
    x = "Monthly Salary (£)",
    y = "Cumulative Probability"
  ) +
  scale_x_continuous(labels = scales::comma_format(prefix = "£")) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

print(cdf_plot)
```

## Salary Percentiles

```{r percentiles}
# Define percentiles of interest
percentiles <- c(0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)

# Calculate salary at each percentile
salary_percentiles <- qlnorm(percentiles, meanlog = mu, sdlog = sigma)

# Create data frame for display
percentile_df <- data.frame(
  Percentile = paste0(percentiles * 100, "%"),
  Monthly_Salary = paste0("£", format(round(salary_percentiles), big.mark=",")),
  Annual_Salary = paste0("£", format(round(salary_percentiles * 12), big.mark=",")),
  Estimated_Employees = round(c(
    num_employees * 0.10,  # 10th percentile
    num_employees * 0.15,  # 25th percentile
    num_employees * 0.25,  # 50th percentile
    num_employees * 0.15,  # 75th percentile
    num_employees * 0.10,  # 90th percentile
    num_employees * 0.05,  # 95th percentile
    num_employees * 0.03   # 99th percentile
  ))
)

# Add executive level
percentile_df <- rbind(percentile_df, data.frame(
  Percentile = "Top (executives)",
  Monthly_Salary = "£22,500+",
  Annual_Salary = "£270,000+",
  Estimated_Employees = 2
))

kable(percentile_df, align = "llrr", caption = "Salary Percentiles and Distribution of Employees")
```

## Simulating Employee Salaries

Let's simulate individual salaries for the 70 employees based on our log-normal distribution.

```{r simulate_employees, fig.width=10, fig.height=6}
# Set seed for reproducibility
set.seed(123)

# Simulate 70 salaries from log-normal distribution
simulated_salaries <- rlnorm(num_employees, meanlog = mu, sdlog = sigma)

# Adjust to ensure total matches exactly £461,000
adjustment_factor <- total_monthly_salary / sum(simulated_salaries)
simulated_salaries <- simulated_salaries * adjustment_factor

# Create data frame for simulated employees
employee_df <- data.frame(
  Employee_ID = 1:num_employees,
  Monthly_Salary = simulated_salaries
) %>%
  arrange(desc(Monthly_Salary))

# Create histogram of simulated salaries
salary_hist <- ggplot(employee_df, aes(x = Monthly_Salary)) +
  geom_histogram(bins = 15, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = mean_salary, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = mean_salary + 1500, y = 10, 
           label = "Mean", color = "red") +
  labs(
    title = "Histogram of Simulated Employee Salaries",
    x = "Monthly Salary (£)",
    y = "Number of Employees"
  ) +
  scale_x_continuous(labels = scales::comma_format(prefix = "£")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

print(salary_hist)

# Display top 10 highest-paid employees
top_10 <- head(employee_df, 10) %>%
  mutate(
    # Store numeric values for calculations
    Monthly_Value = Monthly_Salary,
    Annual_Value = Monthly_Value * 12,
    # Format for display
    Monthly_Salary = paste0("£", format(round(Monthly_Value), big.mark=",")),
    Annual_Salary = paste0("£", format(round(Annual_Value), big.mark=","))
  ) %>%
  select(Employee_ID, Monthly_Salary, Annual_Salary)

kable(top_10, align = "rrr", caption = "Top 10 Highest-Paid Employees (Simulated)")

```

## Comparison with Normal Distribution

To highlight why log-normal is more appropriate than normal distribution for salaries:

```{r normal_comparison, fig.width=10, fig.height=6}
# Calculate normal distribution parameters that would give same mean
normal_mean <- mean_salary
normal_sd <- mean_salary * cv  # Using same coefficient of variation

# Generate normal pdf
normal_pdf <- dnorm(salary_seq, mean = normal_mean, sd = normal_sd)

# Combined data frame for comparison
comparison_df <- data.frame(
  Salary = rep(salary_seq, 2),
  Density = c(pdf_values, normal_pdf),
  Distribution = rep(c("Log-normal", "Normal"), each = length(salary_seq))
)

# Plot comparison
comparison_plot <- ggplot(comparison_df, aes(x = Salary, y = Density, color = Distribution, linetype = Distribution)) +
  geom_line(size = 1) +
  geom_vline(xintercept = mean_salary, linetype = "dashed", color = "black", size = 0.5) +
  annotate("text", x = mean_salary + 1500, y = max(comparison_df$Density) * 0.9, 
           label = "Mean", color = "black") +
  labs(
    title = "Comparison of Log-normal vs Normal Distribution for Salaries",
    x = "Monthly Salary (£)",
    y = "Density"
  ) +
  scale_x_continuous(labels = scales::comma_format(prefix = "£")) +
  scale_color_manual(values = c("Log-normal" = "steelblue", "Normal" = "darkred")) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "top"
  )

print(comparison_plot)
```

## Salary Distribution by Percentile Bands

```{r percentile_bands, fig.width=10, fig.height=6}
# Create bands based on percentiles
bands <- data.frame(
  Band = c("Low (0-25%)", "Lower-Middle (25-50%)", "Upper-Middle (50-75%)", "High (75-90%)", "Very High (90-100%)"),
  Lower = c(0, 0.25, 0.5, 0.75, 0.9),
  Upper = c(0.25, 0.5, 0.75, 0.9, 1),
  Color = c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854")
)

# Calculate salary ranges for each band
bands$Lower_Salary <- qlnorm(bands$Lower, meanlog = mu, sdlog = sigma)
bands$Upper_Salary <- qlnorm(bands$Upper, meanlog = mu, sdlog = sigma)
bands$Midpoint_Salary <- (bands$Lower_Salary + bands$Upper_Salary) / 2
bands$Employee_Count <- round(num_employees * (bands$Upper - bands$Lower))
bands$Total_Band_Salary <- bands$Midpoint_Salary * bands$Employee_Count
bands$Percentage_Budget <- bands$Total_Band_Salary / total_monthly_salary * 100

# Format for display
bands_display <- bands %>%
  select(Band, Lower_Salary, Upper_Salary, Employee_Count, Percentage_Budget) %>%
  mutate(
    Salary_Range = paste0("£", format(round(Lower_Salary), big.mark=","), " - £", 
                        format(round(Upper_Salary), big.mark=",")),
    Percentage_Budget = paste0(round(Percentage_Budget, 1), "%")
  ) %>%
  select(Band, Salary_Range, Employee_Count, Percentage_Budget)

kable(bands_display, align = "llrr", 
      caption = "Salary Distribution by Percentile Bands")

# Create visualization of employee count and budget allocation by band
band_summary <- bands %>%
  select(Band, Employee_Count, Percentage_Budget, Color) %>%
  pivot_longer(cols = c(Employee_Count, Percentage_Budget),
               names_to = "Metric", 
               values_to = "Value")

# Create grouped bar chart
band_plot <- ggplot(band_summary, aes(x = Band, y = Value, fill = Band)) +
  geom_col() +
  facet_wrap(~Metric, scales = "free_y", 
             labeller = labeller(Metric = c("Employee_Count" = "Number of Employees", 
                                           "Percentage_Budget" = "% of Salary Budget"))) +
  scale_fill_manual(values = bands$Color) +
  labs(
    title = "Distribution of Employees and Salary Budget by Band",
    x = "Salary Band",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    legend.position = "none"
  )

print(band_plot)
```

## Lorenz Curve and Gini Coefficient

The Lorenz curve and Gini coefficient are common measures of income inequality:

```{r lorenz_gini, fig.width=10, fig.height=6}
# Sort simulated salaries
sorted_salaries <- sort(simulated_salaries)

# Calculate cumulative percentage of employees and salaries
cum_employees <- (1:num_employees) / num_employees
cum_salaries <- cumsum(sorted_salaries) / sum(sorted_salaries)

# Create data frame for Lorenz curve
lorenz_df <- data.frame(
  cum_employees = c(0, cum_employees),
  cum_salaries = c(0, cum_salaries)
)

# Calculate Gini coefficient
gini <- 1 - 2 * sum(diff(lorenz_df$cum_employees) * 
                    (lorenz_df$cum_salaries[-1] + lorenz_df$cum_salaries[-length(lorenz_df$cum_salaries)])/2)

# Plot Lorenz curve
lorenz_plot <- ggplot(lorenz_df, aes(x = cum_employees, y = cum_salaries)) +
  geom_line(size = 1, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "darkgrey") +
  annotate("text", x = 0.75, y = 0.25, 
           label = paste("Gini Coefficient =", round(gini, 3)), 
           size = 5) +
  labs(
    title = "Lorenz Curve of Salary Distribution",
    x = "Cumulative Proportion of Employees",
    y = "Cumulative Proportion of Total Salary"
  ) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_fixed() +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

print(lorenz_plot)
```

## Conclusion

The log-normal distribution provides a realistic model for company salary structures:

1. The majority of employees fall in the middle salary ranges
2. A small number of highly-paid positions account for a disproportionate percentage of the total salary budget
3. The Gini coefficient of `r round(gini, 3)` indicates a moderate level of inequality, typical for professional organisations

This analysis can help with:
- Budgeting and financial planning
- Comparing compensation structures to industry benchmarks
- Identifying potential compression issues in salary bands
- Developing promotion and compensation strategies

```{r key_metrics_summary}
# Create summary table of key metrics
summary_stats <- data.frame(
  Metric = c(
    "Mean Monthly Salary",
    "Median Monthly Salary",
    "Standard Deviation",
    "Coefficient of Variation",
    "Minimum (10th Percentile)",
    "Maximum (99th Percentile)",
    "Gini Coefficient",
    "Total Monthly Expenditure",
    "Total Annual Expenditure"
  ),
  Value = c(
    paste0("£", format(round(mean_salary), big.mark=",")),
    paste0("£", format(round(exp(mu)), big.mark=",")),
    paste0("£", format(round(mean_salary * cv), big.mark=",")),
    round(cv, 2),
    paste0("£", format(round(qlnorm(0.1, meanlog = mu, sdlog = sigma)), big.mark=",")),
    paste0("£", format(round(qlnorm(0.99, meanlog = mu, sdlog = sigma)), big.mark=",")),
    round(gini, 3),
    paste0("£", format(round(total_monthly_salary), big.mark=",")),
    paste0("£", format(round(total_monthly_salary * 12), big.mark=","))
  )
)

kable(summary_stats, align = "lr", caption = "Summary of Key Salary Distribution Metrics")
```