---
title: "Probabilities in R"
output: html_notebook
---

# Introduction to Probability Functions in R

R provides a suite of functions for working with probability distributions. Each distribution has four variants, specified by a prefix letter:

- `d`: Density/probability mass function (PDF/PMF)
- `p`: Cumulative distribution function (CDF)
- `q`: Quantile function (inverse CDF)
- `r`: Random number generation

For example, for the normal distribution:
- `dnorm()`: PDF value at x
- `pnorm()`: Probability of X ≤ x
- `qnorm()`: Value of x at given probability
- `rnorm()`: Generate random numbers

Let's explore each major distribution:

## 1. Normal Distribution

The normal distribution is defined by its mean (μ) and standard deviation (σ).

```{r}
# Generate random normal numbers
x <- rnorm(1000, mean = 0, sd = 1)
hist(x, breaks = 30, prob = TRUE, main = "Standard Normal Distribution")
curve(dnorm(x), add = TRUE, col = "red")

# PDF at specific points
dnorm(0)  # Density at mean
#> [1] 0.3989423
dnorm(1)  # Density at one SD
#> [1] 0.2419707

# CDF - probability of X ≤ 1.96 (95% confidence interval)
pnorm(1.96)
#> [1] 0.9750021

# Quantile - find value at 97.5th percentile
qnorm(0.975)
#> [1] 1.959964
```

## 2. Binomial Distribution

The binomial distribution models the number of successes in n independent trials with probability p.

```{r}
# Probability of exactly 3 successes in 5 trials with p=0.5
dbinom(3, size = 5, prob = 0.5)
#> [1] 0.3125

# Probability of 3 or fewer successes
pbinom(3, size = 5, prob = 0.5)
#> [1] 0.8125

# Generate 10 random binomial numbers
rbinom(10, size = 5, prob = 0.5)

# Plot probability mass function
x <- 0:10
plot(x, dbinom(x, size = 10, prob = 0.5), type = "h", 
     main = "Binomial Distribution (n=10, p=0.5)")
```

## 3. Poisson Distribution

The Poisson distribution models rare events, with parameter λ representing the average rate.

```{r}
# Probability of exactly 2 events when mean is 1
dpois(2, lambda = 1)
#> [1] 0.1839397

# Probability of 2 or fewer events
ppois(2, lambda = 1)
#> [1] 0.9196986

# Generate random Poisson numbers
rpois(1000, lambda = 2) |> 
  hist(main = "Poisson Distribution (λ=2)")
```

## 4. Uniform Distribution

The uniform distribution models equal probability across an interval.

```{r}
# Generate uniform random numbers
x <- runif(1000, min = 0, max = 1)
hist(x, breaks = 30, prob = TRUE, main = "Uniform Distribution")
curve(dunif(x, 0, 1), add = TRUE, col = "red")

# PDF is constant between min and max
dunif(0.5, min = 0, max = 1)
#> [1] 1

# CDF increases linearly
punif(0.5, min = 0, max = 1)
#> [1] 0.5
```

## 5. Chi-Square Distribution

The chi-square distribution with k degrees of freedom is the sum of k squared standard normal variables.

```{r}
# Generate chi-square random numbers
x <- rchisq(1000, df = 3)
hist(x, breaks = 30, prob = TRUE, main = "Chi-Square Distribution (df=3)")
curve(dchisq(x, df = 3), add = TRUE, col = "red")

# Critical value for 95% confidence
qchisq(0.95, df = 3)
#> [1] 7.814728
```

## 6. Student's t Distribution

The t-distribution is similar to normal but has heavier tails, useful for small samples.

```{r}
# Compare t vs normal densities
curve(dt(x, df = 3), from = -4, to = 4, col = "red",
      main = "t vs Normal Distribution")
curve(dnorm(x), add = TRUE, col = "blue")
legend("topright", c("t(df=3)", "normal"), col = c("red", "blue"), lty = 1)

# Critical value for 95% confidence, df=10
qt(0.975, df = 10)
#> [1] 2.228139
```

## 7. F Distribution

The F distribution is used in ANOVA and regression analysis.

```{r}
# Generate F-distributed random numbers
x <- rf(1000, df1 = 3, df2 = 20)
hist(x, breaks = 30, prob = TRUE, main = "F Distribution")
curve(df(x, df1 = 3, df2 = 20), add = TRUE, col = "red")

# Critical value for 95% confidence
qf(0.95, df1 = 3, df2 = 20)
#> [1] 3.098391
```

## Common Use Cases

1. Normal Distribution:
   - Confidence intervals
   - Statistical inference
   - Modeling natural phenomena

2. Binomial Distribution:
   - A/B testing
   - Quality control
   - Success/failure experiments

3. Poisson Distribution:
   - Modeling rare events
   - Queue theory
   - Count data

4. Uniform Distribution:
   - Random number generation
   - Simulation studies
   - Testing algorithms

5. Chi-Square Distribution:
   - Goodness of fit tests
   - Tests of independence
   - Variance analysis

6. Student's t Distribution:
   - Small sample inference
   - Confidence intervals
   - Hypothesis testing

7. F Distribution:
   - ANOVA
   - Model comparison
   - Variance ratio tests

## Practical Tips

1. Always check your parameters:
```{r}
# This will give a warning
rnorm(10, sd = -1)
```

2. Use `set.seed()` for reproducible results:
```{r}
set.seed(123)
rnorm(3)
set.seed(123)
rnorm(3)  # Same numbers
```

3. For probabilities, remember:
   - `d` functions return height of PDF/PMF
   - `p` functions return area (probability)
   - `q` functions are inverse of `p`
   - `r` functions generate random numbers