---
title: "Understanding R Syntax: A Beginner's Guide"
author: "Nathan Ormond"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
```

# Introduction to R Syntax

This guide explains the fundamental syntax elements of R, including how to create objects, work with data structures, access elements using the `$` syntax, and call functions. Understanding these basics will help you navigate R code and write your own scripts more effectively.

## Creating Objects in R

In R, you create objects by assigning values to names using the assignment operator `<-` (or sometimes `=`).

```{r basic_assignment}
# Create a simple numeric object
x <- 42

# Create a character (string) object
name <- "John Smith"

# Print the objects to see their values
x
name

# You can also use = for assignment, but <- is the convention in R
y = 100
y
```

The convention in R is to use `<-` for assignments rather than `=`, though both work in most contexts. The assignment operator can be read as "gets" - so `x <- 42` reads as "x gets 42".

## Basic Data Types in R

R has several fundamental data types:

```{r data_types}
# Numeric (includes integers and floating-point)
num <- 42.5
typeof(num)

# Integer (add L to specify an integer explicitly)
int <- 42L
typeof(int)

# Character (strings)
text <- "Hello, R!"
typeof(text)

# Logical (TRUE/FALSE)
logic <- TRUE
typeof(logic)

# Complex numbers
comp <- 3 + 4i
typeof(comp)
```

## Creating Different Data Structures

R has several primary data structures that you'll use frequently:

### Vectors

Vectors are one-dimensional arrays that can hold elements of the same type:

```{r vectors}
# Create a numeric vector
numbers <- c(10, 20, 30, 40, 50)
numbers

# Create a character vector
fruits <- c("apple", "banana", "cherry")
fruits

# Create a logical vector
is_approved <- c(TRUE, FALSE, TRUE, TRUE)
is_approved

# Check the length of a vector
length(numbers)

# Vectors can only hold one data type - they will convert if mixed
mixed <- c(1, 2, "three", 4)
mixed  # All elements became characters
```

### Matrices

Matrices are two-dimensional arrays with rows and columns, holding elements of the same type:

```{r matrices}
# Create a 3x3 matrix
mat <- matrix(1:9, nrow = 3, ncol = 3)
mat

# You can fill matrices by row (default is by column)
mat_by_row <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)
mat_by_row

# Get dimensions of a matrix
dim(mat)
```

### Lists

Lists can contain elements of different types and even other lists:

```{r lists}
# Create a list with mixed data types
my_list <- list(
  name = "Alice",
  age = 30,
  scores = c(85, 90, 78),
  passed = TRUE
)
my_list

# Lists can contain other lists (nested structure)
nested_list <- list(
  person = list(name = "Bob", age = 25),
  data = c(1, 2, 3)
)
nested_list
```

### Data Frames

Data frames are table-like structures for storing data in rows and columns, similar to spreadsheets:

```{r dataframes}
# Create a data frame
students <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(22, 25, 23),
  grade = c("A", "B", "A-"),
  stringsAsFactors = FALSE  # Prevents automatic conversion of strings to factors
)
students

# See the structure of a data frame
str(students)

# Get dimensions (rows and columns)
dim(students)

# Get column names
names(students)
```

## Using the `$` Syntax for Accessing Elements

The `$` operator is used to access elements in lists and data frames:

### Accessing List Elements with `$`

```{r dollar_lists}
# Access elements of a list using $
my_list$name
my_list$scores

# Access nested elements with multiple $
nested_list$person$name
```

### Accessing Data Frame Columns with `$`

```{r dollar_dataframes}
# Access columns in a data frame
students$name
students$age

# You can also create or modify columns using $
students$passed <- c(TRUE, TRUE, FALSE)
students
```

## Alternative Ways to Access Elements

Besides the `$` operator, there are other ways to access elements:

### Using Square Brackets `[ ]`

```{r square_brackets}
# For vectors: access by position
numbers[3]  # Get the third element
numbers[c(1, 3, 5)]  # Get the first, third, and fifth elements

# For matrices: access by row and column [row, column]
mat[2, 3]  # Element at second row, third column
mat[1, ]   # Entire first row
mat[, 2]   # Entire second column

# For data frames: similar to matrices
students[2, 3]  # Value at second row, third column
students[, "age"]  # Entire age column
students[students$age > 23, ]  # Rows where age > 23
```

### Using Double Brackets `[[ ]]` for Lists

```{r double_brackets}
# Access list elements by position
my_list[[1]]  # First element
my_list[[3]]  # Third element (scores vector)

# You can also use names with [[ ]]
my_list[["name"]]
```

## Working with Functions in R

Functions in R take inputs (arguments) and produce outputs:

### Calling Basic Functions

```{r basic_functions}
# Mathematical functions
sqrt(25)
log(10)
abs(-15)

# Statistical functions
mean(c(10, 20, 30, 40, 50))
median(c(10, 20, 30, 40, 50))
sd(c(10, 20, 30, 40, 50))
```

### Function Arguments

Functions can have multiple arguments:

```{r function_arguments}
# Named arguments
round(3.14159, digits = 2)

# Position-based arguments (same result)
round(3.14159, 2)

# Default arguments
seq(from = 1, to = 10)  # 1, 2, 3, ..., 10

# Changing default arguments
seq(from = 1, to = 10, by = 2)  # 1, 3, 5, 7, 9
```

### Creating Your Own Functions

You can define your own functions:

```{r custom_functions}
# Define a simple function
square <- function(x) {
  return(x^2)
}

# Use your function
square(5)
square(1:5)  # Works on vectors too!

# Function with multiple arguments
calculate_bmi <- function(weight, height) {
  bmi <- weight / (height^2)
  return(bmi)
}

calculate_bmi(weight = 70, height = 1.75)
```

## Practical Example: Putting It All Together

Let's combine the concepts we've learned in a practical example:

```{r practical_example}
# Create a data frame with exam scores
exam_data <- data.frame(
  student_id = c(101, 102, 103, 104, 105),
  name = c("Alice", "Bob", "Charlie", "David", "Emma"),
  math = c(85, 76, 92, 68, 94),
  english = c(90, 88, 75, 82, 89),
  science = c(78, 84, 90, 79, 93)
)

# Display the data
exam_data

# Calculate the average score for each student
exam_data$average <- (exam_data$math + exam_data$english + exam_data$science) / 3
exam_data

# Create a passing grade column (TRUE if average >= 80)
exam_data$passed <- exam_data$average >= 80
exam_data

# Function to determine letter grade
assign_grade <- function(score) {
  if (score >= 90) {
    return("A")
  } else if (score >= 80) {
    return("B")
  } else if (score >= 70) {
    return("C")
  } else if (score >= 60) {
    return("D")
  } else {
    return("F")
  }
}

# Apply the function to each student's average (using sapply)
exam_data$letter_grade <- sapply(exam_data$average, assign_grade)
exam_data

# Get summary statistics for math scores
summary(exam_data$math)

# Find students who excelled in all subjects (scored at least 80 in each)
excelled <- exam_data[exam_data$math >= 80 & exam_data$english >= 80 & exam_data$science >= 80, ]
excelled
```

## Common R Operations and Syntax Pattern

### Logical Operators

```{r logical_operators}
# Equality
5 == 5

# Inequality
5 != 4

# Greater than/less than
5 > 3
5 < 10

# AND operator
(5 > 3) & (4 < 7)

# OR operator
(5 > 10) | (4 < 7)

# NOT operator
!(5 == 3)
```

### Control Structures

```{r control_structures}
# If-else statement
x <- 42
if (x > 50) {
  print("x is greater than 50")
} else if (x > 40) {
  print("x is between 41 and 50")
} else {
  print("x is 40 or less")
}

# For loop
for (i in 1:5) {
  print(paste("Iteration", i, ":", i^2))
}

# While loop
counter <- 1
while (counter <= 5) {
  print(paste("Counter:", counter))
  counter <- counter + 1
}
```

### Using the Pipe Operator (`%>%`)

The pipe operator (`%>%`) from the `magrittr` package (loaded by `dplyr`) allows for more readable code by passing the result of one function to another:

```{r pipe_operator}
# Load dplyr (for pipe operator and data manipulation)
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr", quiet = TRUE)
}
library(dplyr)

# Without pipe operator
head(arrange(filter(exam_data, math > 80), desc(average)))

# With pipe operator - much more readable!
exam_data %>% 
  filter(math > 80) %>%
  arrange(desc(average)) %>%
  head()
```

## Summary of R Syntax Patterns

Here's a quick reference for the syntax patterns we've covered:

1. **Assignment**: `x <- value`
2. **Vector creation**: `c(value1, value2, ...)`
3. **Accessing elements**:
   - Lists/data frames with `$`: `object$element_name`
   - Single elements with `[ ]`: `object[index]` or `object[row, column]`
   - List elements with `[[ ]]`: `list[[index]]` or `list[["name"]]`
4. **Function calls**: `function_name(arg1, arg2, ...)`
5. **Creating functions**: 
   ```r
   function_name <- function(arg1, arg2) {
     # code
     return(result)
   }
   ```
6. **Logical operations**: `&` (AND), `|` (OR), `!` (NOT), `==` (equals), `!=` (not equals)
7. **Piping operations**: `data %>% function1() %>% function2()`

Understanding these syntax patterns will help you read and write R code effectively and efficiently.

## Exercises for Practice

Try these exercises to reinforce your understanding of R syntax:

1. Create a vector of the first 10 even numbers.
2. Create a data frame with 3 columns: "product", "price", and "quantity".
3. Calculate the total cost for each product (price × quantity) and add it as a new column.
4. Create a function that converts temperatures from Celsius to Fahrenheit.
5. Use the function to convert a vector of temperatures.